---
title: "Tool_Setup"
format: html
editor: visual
---

## Overview

This tutorial walks the user through the initial setup of the Forecasting Lake And Reservoir Ecosystems (FLARE) user tool. This tutorial instructs the user on installing the necessary packages, setting up the .yml, checking for remote sensing data availability, downloading bathymetry, estimating light extinction coefficient, and running a short FLARE forecast.

Both this tutorial file and other code in the repository are best run in a Docker container. The README file contains instructions for a Mac user to run this code in a Docker.

After finishing this setup and confirming data availability, the user may go to Run_Tool.R to run FLARE more quickly and, if desired, iteratively.

### Install packages

If you are running this in the provided Docker (recommended), some packages are already installed. If you are running this outside of the contained, you will need to install all packages.

```{r}
# remotes::install_github('rqthomas/GLM3r')
# remotes::install_github("FLARE-forecast/FLAREr")
install.packages('pacman')
pacman::p_load('rstac', 'terra', 'stars',
               'ggplot2', 'tidyterra', 'viridis',
               'yaml','gdalcubes', 'tmap', 'dplyr',
               'tidyverse', 'sf','arrow', 'raster',
               'elevatr', 'marmap', 'rLakeAnalyzer',
               'httr', 'jsonlite', 'readr')
remotes::install_github('usgs-r/glmtools', force = T, upgrade = 'never')
remotes::install_github("FLARE-forecast/ropenmeteo", force = T, upgrade = "never")
library(glmtools)
library(ropenmeteo)
library(FLAREr)
```

### Set up yml file

A .yml is a plain-text file used for storing and organizing configuration data. Here, the user inputs their desired lake information in configure_rs_run.yml. To do this, open the configure_rs_run.yml file.

In the file, the user needs to input multiple pieces of information.

1.  Input both a start_datetime (the date on which the spin-up period starts) and a forecast_start_datetime (the date on which the forecast itself begins) in the format YYYY-MM-DD HH:MM:SS format.
2.  Input your desired forecast horizon
3.  Input a four-letter abbreviation of the lake name of interest (for example, 'wald' for Walden Pond)
4.  Input the bounding box coordinates of your lake of interest
5.  Input a 'focal point', or representative point, of your lake of interest. For example, if your lake is a perfect circle, the middle of the circle would be an appropriate point.
6.  Under "Kw", set type to either type: us if your lake is in the United States, or type: global if your lake is outside of the United States.
7.  Optional: If you know the maximum depth of your lake, input this value to max_depth_allowed to better constrain the bathymetry estimates.

### Get info from yml to grab data

```{r}
# This need to be set to run each experiment
run_name <- "run"
config_flare_file <- "configure_flare.yml"
config_set_name <- "analysis"
configure_run_file <- "configure_run.yml"
use_s3 <- FALSE
lake_directory <- here::here()
# Point FLARE to GLM location
# Sys.setenv('GLM_PATH' = paste0(lake_directory, '/binary/macos/glm')) # if running on local computer
Sys.setenv('GLM_PATH' = '/glm') # if running in FLARE rocker
options(future.globals.maxSize = 891289600)
# Get scripts needed to run this code
inputs <- read_yaml("configure_rs_run.yml")
walk(list.files(file.path(here::here(), "R"), pattern = "\\.R$", full.names = TRUE), source)
# Get data from yml
num_lakes <- length(inputs$lakes)
lake_num <- 1
start_date <- as_date(inputs$start_datetime)
forecast_start_date <- as_date(inputs$forecast_start_datetime)
forecast_end_date <- as_date(inputs$start_datetime) + days(inputs$forecast_horizon)
site_id <- names(inputs$lakes[lake_num])
bbox_list <- inputs$lakes[lake_num][[1]][[1]]
bbox <- c(left = bbox_list$left,
          bottom = bbox_list$bottom,
          right = bbox_list$right,
          top = bbox_list$top)
focal_point_list <- inputs$lakes[lake_num][[1]][[2]]
kw_list <- inputs$lakes[lake_num][[1]][[3]]
points_df <- data.frame(lon = c(focal_point_list$lon), lat = focal_point_list$lat)
points <- st_as_sf(x = points_df,
                   coords = c("lon", "lat"),
                   crs = 4326)
max_depth_allowed <- inputs$lakes[lake_num][[1]][[4]]
```

### Download remote sensing temperature data

This will pull thermal remote sensing imagery (Landsat 8/9 Surface Temperature Product) from a Spatiotemporal Asset Catalog (STAC) and save out as a targets file for FLARE. Depending on your selected dates, there may be no cloud-free imagery available. If this is the case, you will get a message stating this and may need to select different dates, or a longer date range, in your .yml. Similarly, if you have selected a very long date range (more than a couple months), this may take a long time to run and download data.

```{r}
thermaldata <- get_lst(bbox,
                       start_date = format(as_datetime(inputs$start_date), "%Y-%m-%dT%H:%M:00Z"),
                       end_date = format(as_datetime(inputs$forecast_start_datetime) + days(inputs$forecast_horizon),"%Y-%m-%dT%H:%M:00Z"),
                       points,
                       spinup_start = format(as_datetime(inputs$start_datetime), "%Y-%m-%dT%H:%M:00Z"),
                       spinup_end = format(as_datetime(inputs$forecast_start_datetime), "%Y-%m-%dT%H:%M:00Z"))
# see what it looks like!
ggplot() +
  geom_stars(data = thermaldata["thermal_C"]) +
  facet_wrap(~time) +
  theme_classic() +
  scale_fill_viridis(na.value = 'transparent') +
  labs(fill = "Temperature (C)") +
  coord_fixed()
# get values
thermal_vals <- get_vals(points, thermaldata)
output <- clean_data(thermal_vals, site_id)

# create directory for targets file & save
dir.create(paste0('./targets/', site_id, '/'), recursive = T)
write_csv(output, paste0('targets/', site_id, '/', site_id, '-targets-rs.csv'))
```

### Download remote sensing height data

This will download water surface elevation data from the SWOT satellite.

```{r}
swot_data <- get_swot(bbox, 
                      start_date = as.Date(inputs$start_date),
                      end_date = as.Date(inputs$forecast_start_datetime) + days(inputs$forecast_horizon),
                      site_id)
if(nrow(swot_data) != 0){
 swot_data$datetime <- as.Date(swot_data$datetime)
}
targets <- read_csv(paste0('targets/', site_id, '/', site_id, '-targets-rs.csv'))
targets <- rbind(swot_data, targets)
write_csv(targets, paste0('targets/', site_id, '/', site_id, '-targets-rs.csv'))

```

### Download local meteorology

This downloads stage 2 and stage 3 meteorology data from dynamical.org. Depending on your date range, this may take a while. Python must be installed to run this.

```{r}
start_date <- as_date(inputs$start_datetime)
forecast_start_date <- as_date(inputs$forecast_start_datetime)
forecast_end_date <- as_date(inputs$start_datetime) + days(inputs$forecast_horizon)

# stage 3
.run_get_met_py(
  stage = "stage3",
  site_id = site_id,
  bbox = bbox,
  reference_datetime = start_date,
  end_date = forecast_start_date
)

# stage 2
.run_get_met_py(
  stage = "stage2",
  site_id = site_id,
  bbox = bbox,
  reference_datetime = forecast_start_date
)
```

### Get bathymetry

This downloads bathymetric data from the GLOBathy dataset. Confirm that the bathymetry is for the correct lake by looking at the output plot. If the bathymetry looks too deep, set max_depth_allowed in the .yml. If the user has their own bathymetry raster or existing H/A relationship, they may use this instead.

```{r}
bathy <- find_matches(bbox)
plot(bathy) # visually check this is the correct lake
ha <- get_ha(bathy, points) # user may input their own raster instead of 'bathy' or set ha <- 'user H/A'

if(!is.na(max_depth_allowed) & !is.null(max_depth_allowed)){
  abs_z = ha$depths[1] - ha$depths
  ha <- ha[which(abs_z <= max_depth_allowed), ]
}

bsn_dims <- get_bsn_dims(bathy)
```

### Get light extinction coefficient (Kw)

The light extinction coefficient is estimated from Secchi depth data. The US dataset (using LAGOS US) is much more extensive than the global dataset. If the user's lake does not exist in these datasets, they may manually set Kw in the .yml based on general knowledge about the turbidity of the lake.

```{r}
# Search for lake in LAGOS US database
if(kw_list$type == "us"){
  if(is.na(kw_list$lake_row) || is.null(kw_list$lake_row)){
    kw_list$lake_row <- 1
  }
  mylake_kw <- get_kw_US(bbox, validate = FALSE, lake_row = kw_list$lake_row)
}else if(kw_list$type == "global"){
  mylake_kw <- get_kw_global(bbox, validate = FALSE)
}else{
  if(!is.na(kw_list$value) & !is.null(kw_list$value)){
    mylake_kw <- kw_list$value
  }else{
    stop("missing user specified kw value")
  }
}
```

### Update FLARE configuration files

Based on the data downloaded above, this code chunk updates the GLM and configuration files for FLARE.

```{r}
yml <- yaml::read_yaml(file.path("configuration", config_set_name, config_flare_file))
yml$location$site_id <- site_id
yml$location$latitude <- points_df[[2]][1]
yml$location$longitude <- points_df[[1]][1]
yml$default_init$lake_depth <- rev(max(ha$depths) - min(ha$depths))
modeled_depths <- seq(0, yml$default_init$lake_depth-1)
#yml$default_init$temp <- rep(sed_data$water_temp_init[1], times = length(modeled_depths))
#yml$default_init$temp_depths <- modeled_depths
yml$model_settings$modeled_depths <- modeled_depths
yml$da_setup$obs_filename <- paste0(site_id, "-targets-rs.csv")

yaml::write_yaml(yml, file.path("configuration", config_set_name, config_flare_file))

# create list of variable values & names for input to nml
var_list <- list(site_id, mylake_kw, site_id, points_df[[2]][1], points_df[[1]][1],
                 dim(ha)[1], rev(ha$depths), rev(ha$Area.at.z), rev(max(ha$depths) - min(ha$depths))-0.5,
                 bsn_dims$bsn_len, bsn_dims$bsn_wid)
var_name_list <- list("sim_name", "Kw", "lake_name", "latitude", "longitude",
                      "bsn_vals", "H", "A", "lake_depth",
                      "bsn_len", "bsn_wid")

# update nml
update_nml(var_list, var_name_list,
           working_directory = file.path("configuration", config_set_name), nml = 'glm3.nml')

run_config <- yaml::read_yaml(file.path("configuration", config_set_name, configure_run_file))
run_config$start_datetime <- inputs$start_datetime
run_config$forecast_horizon <- inputs$forecast_horizon
run_config$forecast_start_datetime <- inputs$forecast_start_datetime
run_config$sim_name <- run_name

dir.create(file.path("./restart", site_id, run_name), recursive = T)

yaml::write_yaml(run_config, file = file.path("restart", site_id, run_name, configure_run_file))
```

### Run FLARE

This line will run a FLARE forecast. The output plots may be found in /plots/YOUR LAKE NAME.

```{r}
# If you get an error here, you may need to install the latest version of FLARE
# remotes::install_github("FLARE-forecast/FLAREr")

FLAREr::run_flare(lake_directory, configure_run_file = configure_run_file, config_set_name = config_set_name,
                  clean_start = FALSE, sim_name = run_name)
```
